---
title: "Merge Author Data"
author: "Ben Schmidt"
date: "February 24, 2015"
output: html_document
---

How do you merge in author data to our set?

I've added a file to the server that contains labels for library of congress classifications.

```{r}
library(dplyr)
classifications = read.table("data/LOC_classification.tsv",sep="\t",header=T) %>% as.tbl
classifications %>% glimpse
```

This wants to work with our existing title data, which also has a `lc1` field.

``` {r}
shortTitles = read.table("data/booktitlesShort.tsv",sep="\t",quote="",comment.char = "")

head(shortTitles)
```

This can be done with the `inner_join` function of dplyr, which by default merges two tables based on their shared column names.

``` {r}

labelled = shortTitles %>% inner_join(classifications)

```

Note that `inner_join` drops rows that aren't contained in both tables. Only about a third of the books in the `shortTitles` folder have an lc1 that matches the classifications field. The `nrow` function tells how many rows a datafraem has: here you can see that joining it to the genre lookup loses almost two-thirds the entries.

``` {r}
nrow(shortTitles)
nrow(labelled)

```

To see what's being dropped, you can use the `anti_join` function that shows only elements in the first frame that are unmatched in the second one.


``` {r}
shortTitles %>% anti_join(classifications) %>% sample_n(30) %>% select(lc1)
```

If you want to keep all elements from the left frame and just put `NA` values into the various fields defined by the right one, you can use the `left_join` function which prioritizes things from the left. (`outer_join` privileges both fields, and `right_join` privileges the second one.)

``` {r}
shortTitles %>% left_join(classifications) %>% nrow
```
